---
title: "Explorartory Data Analysis for Heart Faliure Clinical Records "
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2023-11-29"

Dataset :
Heart failure clinical records. (2020).
UCI Machine Learning Repository. https://doi.org/10.24432/C5Z89R.
---

## Contents

1.  Summary of data

2.  Explore Factors effecting Mortality

3.  Summary Statistics for Dependent Features

4.  Visualization

5.  Future Considerations

## 1. Summary of Data

```{r}

# Create a copy of the dataset
dataset <- read.csv("heart_failure_clinical_records_dataset.csv")


dataset_new <- dataset

# Loop through each column in the dataset
for (col_name in names(dataset_new)) {
  # Check if the column contains only 0 and 1
  if (all(dataset_new[[col_name]] %in% c(0, 1))) {
    # Convert the column to a factor with labels 'No' for 0 and 'Yes' for 1
    dataset_new[[col_name]] <- factor(dataset_new[[col_name]], levels = c(0, 1), labels = c('No', 'Yes'))
  }
}

```

```{r}
# Assuming 'dataset_new' is your data frame

# Initialize an empty data frame to store the results
summary_table <- data.frame(Column = character(), Type = character(), Range = character(), stringsAsFactors = FALSE)

# Loop through each column in the dataset_new
for (col_name in names(dataset_new)) {
  # Determine the type of column
  col_type <- class(dataset_new[[col_name]])
  
  # Initialize the range variable
  col_range <- NA
  
  # If the column is numeric or integer, get the range
  if (col_type %in% c('numeric', 'integer')) {
    col_range <- paste(min(dataset_new[[col_name]], na.rm = TRUE), max(dataset_new[[col_name]], na.rm = TRUE), sep = " - ")
  }
  
  # If the column is a boolean (logical), identify as Boolean and set range as TRUE/FALSE
  if (col_type == 'logical') {
    col_type <- 'Boolean'
    col_range <- "FALSE - TRUE"
  }
  
  # If the column is already a factor, identify as Categorical
  if (col_type == 'factor') {
    col_type <- 'Categorical'
    col_range <- paste(levels(dataset_new[[col_name]]), collapse = ", ")
  }
  
  # Add the information to the summary table
  summary_table <- rbind(summary_table, data.frame(Column = col_name, Type = col_type, Range = col_range))
}

# Print the summary table
print(summary_table)

# Write the summary table to a CSV file
write.csv(summary_table, file = "column_summary_table.csv", row.names = FALSE)


```

```{r}
# Initialize an empty data frame to store the normality results
normality_table <- data.frame(
  Column = character(), 
  Shapiro_Wilk_p_value = numeric(), 
  Normality = character(), 
  stringsAsFactors = FALSE
)

# Define the threshold for normality
normality_threshold <- 0.05

# Loop through each column in the dataset_new
for (col_name in names(dataset_new)) {
  # Determine the type of column
  col_type <- class(dataset_new[[col_name]])
  
  # Initialize the p-value and normality variables
  shapiro_p_value <- NA
  normality_status <- "Not Applicable"

  # If the column is numeric or integer, perform Shapiro-Wilk test
  if (col_type %in% c('numeric', 'integer')) {
    shapiro_test <- shapiro.test(dataset_new[[col_name]])
    shapiro_p_value <- shapiro_test$p.value
    normality_status <- ifelse(shapiro_p_value > normality_threshold, "Normal", "Not Normal")
  }
  
  # Add the normality information to the normality table
  normality_table <- rbind(normality_table, data.frame(
    Column = col_name, 
    Shapiro_Wilk_p_value = shapiro_p_value, 
    Normality = normality_status
  ))
}

# Print the normality table
normality_table_False <- normality_table %>% filter(Normality =="Not Normal")
print(normality_table_False)

# Write the normality table to a CSV file
write.csv(normality_table, file = "normality_assessment_table.csv", row.names = FALSE)

```

## 2. Feature Effect on Mortality

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Initialize an empty data frame to store the results
mortality_features <- data.frame(
  Column = character(), 
  Test = character(), 
  P_Value = numeric(), 
  Statistical_Significance = logical(), 
  stringsAsFactors = FALSE
)

# Loop through each column, excluding DEATH_EVENT
for (col_name in setdiff(names(dataset_new), "DEATH_EVENT")) {
  test <- ""
  p_value <- NA
  significance <- NA

  # Check if the column is a factor and convert it to character for analysis
  if (is.factor(dataset_new[[col_name]])) {
    dataset_new[[col_name]] <- as.character(dataset_new[[col_name]])
  }

  # Determine if the column is categorical or numerical
  if (is.character(dataset_new[[col_name]]) || is.factor(dataset_new[[col_name]])) {
    # Chi-Square test for categorical variables
    test <- "Chi-Square"
    tbl <- table(dataset_new[[col_name]], dataset_new$DEATH_EVENT)
    p_value <- chisq.test(tbl)$p.value
  } else {
    # Shapiro-Wilk test for normality
    shapiro_test <- shapiro.test(dataset_new[[col_name]])
    
    if (shapiro_test$p.value > 0.05) {
      # Data is normally distributed, use T-Test
      test <- "T-Test"
      p_value <- t.test(dataset_new[[col_name]] ~ dataset_new$DEATH_EVENT)$p.value
    } else {
      # Data is not normally distributed, use Mann-Whitney U Test
      test <- "Mann-Whitney U"
      p_value <- wilcox.test(dataset_new[[col_name]] ~ dataset_new$DEATH_EVENT)$p.value
    }
  }

  # Determine if dependent based on p-value
  significance<- p_value < 0.05

  # Add results to the data frame
  mortality_features <- rbind(mortality_features, data.frame(
    Column = col_name, 
    Test = test, 
    P_Value = p_value, 
    Statistical_Significance = significance
  ))
}

# Write the results to CSV files
mortality_features_True <- mortality_features %>% filter(Statistical_Significance == TRUE)
write.csv(mortality_features, file = "mortality_features_analysis_with_DEATH_EVENT.csv", row.names = FALSE)
write.csv(mortality_features_True, file = "mortality_features_analysis_with_DEATH_EVENT_True.csv", row.names = FALSE)
print(mortality_features_True)

```

## 2.1 Question: How do multiple clinical factors simultaneously impact the risk of death in heart failure patients?

Target Conclusion: Identify key predictors of mortality and quantify their impact while controlling for other variables.

```{r}
# Assuming 'dataset_new' is your data frame, 'DEATH_EVENT' is the binary outcome variable,
# and you have several clinical factors like age, blood pressure, etc.

# Fit logistic regression model
model <- glm(DEATH_EVENT ~ age + ejection_fraction + serum_creatinine + serum_sodium + diabetes +smoking, 
             family = binomial(link = "logit"), data = dataset_new)

# Summarize the model to see the coefficients and significance
summary(model)

# To get the odds ratio for better interpretability
exp(coef(model))

```

Based on the output of the logistic regression model:

-   **Age**: The positive coefficient suggests that as age increases, so does the risk of death. It is statistically significant (p \< 0.001), indicating a reliable predictor.
-   **Ejection Fraction**: The negative coefficient indicates that higher qz¸Œ!q ejection fraction values are associated with a lower risk of death, and this predictor is also statistically significant (p \< 0.001).
-   **Serum Creatinine**: This has a positive coefficient and is statistically significant (p \< 0.001), suggesting that higher serum creatinine levels increase the risk of death.
-   **Serum Sodium**: The coefficient is negative, but it is not statistically significant (p \> 0.05), implying that serum sodium levels may not be a reliable predictor of death risk in this model.
-   **Diabetes**: The positive coefficient is not statistically significant (p \> 0.05), suggesting that diabetes, as coded, may not be associated with the risk of death in this model.
-   **Smoking**: The negative coefficient is not statistically significant (p \> 0.05), indicating that smoking status may not be a significant predictor of death risk.

The significant predictors of mortality risk in this cohort of heart failure patients are age, ejection fraction, and serum creatinine levels.

Each increase in these parameters (age and serum creatinine) is associated with an increased risk of death, while higher ejection fraction is associated with decreased risk.

## 3. Summary Statistics for Dependent Features

```{r}
library(dplyr)

# Group the dataset by Death Event
summary_table <- dataset %>%
  group_by(DEATH_EVENT) %>%
  summarize(
    Mean_Age = mean(age),
    SD_Age = sd(age),
    Median_Age = median(age),
    Mean_Ejection_Fraction = mean(ejection_fraction),
    SD_Ejection_Fraction = sd(ejection_fraction),
    Median_Ejection_Fraction = median(ejection_fraction),
    Mean_Serum_Creatinine = mean(serum_creatinine),
    SD_Serum_Creatinine = sd(serum_creatinine),
    Median_Serum_Creatinine = median(serum_creatinine)
  )

# Transpose the summary table
summary_table <- t(summary_table)
 


# Print the table
print(table)

```

Age: Patients who experienced death (DEATH_EVENT=1) tend to be older on average.

Ejection Fraction: Patients who survived (DEATH_EVENT=0) have a higher average ejection fraction, indicating better heart function.

Serum Creatinine: Patients who did not survive have higher average serum creatinine levels, which may indicate kidney issues.

## 4.Graphs

1.  **Box Plot of Serum Creatinine by Death Event**:
    -   People who experienced a death event tend to have higher levels of serum creatinine.
2.  **Box Plot of Ejection Fraction by Death Event**:
    -   Lower ejection fractions are more commonly associated with death events.
3.  **Box Plot of Age by Death Event**:
    -   Those who experienced a death event are generally older.
4.  **Correlation Heatmap**:
    -   The heatmap indicates a stronger relationship between serum creatinine and ejection fraction than with age or serum sodium.
5.  **Serum Creatinine vs. Death Event Bar Chart**:
    -   A larger proportion of individuals with lower serum creatinine levels did not experience a death event.
6.  **Ejection Fraction vs. Death Event Bar Chart**:
    -   The majority of individuals with higher ejection fractions did not experience a death event.
7.  **Age vs. Death Event Bar Chart**:
    -   There are more non-death events across all age groups, but the age group with the highest number of death events is also the largest group.
8.  **High Blood Pressure vs. Death Event Bar Chart**:
    -   High blood pressure doesn't show a strong difference in the number of death events versus non-death events.
9.  **Diabetes vs. Death Event Bar Chart**:
    -   Having diabetes does not show a significant difference in the occurrence of death events.
10. **Smoking vs. Death Event Bar Chart**:
    -   Non-smokers have a higher count of non-death events, but the difference between smokers and non-smokers regarding death events is not pronounced.

These conclusions suggest that higher serum creatinine levels and lower ejection fractions are associated with a higher occurrence of death events, whereas high blood pressure, diabetes, and smoking status do not show a clear association with death events. Age shows a trend where older individuals are more likely to have experienced a death event.

11. **Scatter Plot of Age vs. Ejection Fraction**:

-   This scatter plot suggests no strong correlation between age and ejection fraction for those who did not experience a death event, as the points are spread throughout the plot. For those with a death event, while also spread, there's a slight indication of higher age and lower ejection fraction occurrences together.

12. **Scatter Plot of Serum Creatinine vs. Ejection Fraction**:

-   For individuals without a death event, serum creatinine levels are generally lower regardless of ejection fraction, while for those with a death event, higher serum creatinine levels are more common, especially as ejection fractions decrease.

These plots imply that among individuals who experienced a death event, there is a tendency towards higher serum creatinine levels and lower ejection fractions, while age does not show a distinct pattern in relation to ejection fraction.

```{r}

library(ggplot2)

# Create the scatter plot with facets
scatter_plot <- ggplot(data = dataset_new, aes(x = ejection_fraction, y = age, color = factor(DEATH_EVENT))) +
  geom_point() + scale_color_manual(values = c("blue", "red")) +
  labs(
    title = "Scatter Plot of Age vs. Ejection Fraction",
    x = "Ejection Fraction",
    y = "Age",
    color = "Death Event"
  ) +
  facet_wrap(~DEATH_EVENT)

# Show the scatter plot
print(scatter_plot)



library(ggplot2)

# Create the scatter plot with facets
scatter_plot <- ggplot(data = dataset_new, aes(x = ejection_fraction, y = serum_creatinine, color = factor(DEATH_EVENT))) +
  geom_point() + scale_color_manual(values = c("blue", "red")) +
  labs(
    title = "Scatter Plot of Serum Creatinine vs. Ejection Fraction",
    x = "Ejection Fraction",
    y = "Serum Creatinine,",
    color = "Death Event"
  ) +
  facet_wrap(~DEATH_EVENT)

# Show the scatter plot
print(scatter_plot)

```

```{r}

# Create bar charts for categorical variables
bar_chart_hp <- ggplot(data = dataset_new, aes(x = high_blood_pressure, fill = factor(DEATH_EVENT))) +
  geom_bar(position = "dodge") +  scale_fill_manual(values = c("blue", "red"))+
  labs(
    title = "High Blood Pressure vs. Death Event",
    x = "High Blood Pressure",
    y = "Count",
    fill = "Death Event"
  )

bar_chart_diabetes <- ggplot(data = dataset_new, aes(x = diabetes, fill = factor(DEATH_EVENT))) +
  geom_bar(position = "dodge") + scale_fill_manual(values = c("blue", "red"))+
  labs(
    title = "Diabetes vs. Death Event",
    x = "Diabetes",
    y = "Count",
    fill = "Death Event"
  )

bar_chart_smoking <- ggplot(data = dataset_new, aes(x = smoking, fill = factor(DEATH_EVENT))) +
  geom_bar(position = "dodge") +   scale_fill_manual(values = c("blue", "red"))+
  labs(
    title = "Smoking vs. Death Event",
    x = "Smoking",
    y = "Count",
    fill = "Death Event"
  )

# Show the bar charts
print(bar_chart_hp)
print(bar_chart_diabetes)
print(bar_chart_smoking)

```

```{r}

# Create age bins
dataset_new$age_group <- cut(dataset_new$age, breaks = c(39, 50, 60, 70, 80,90, 100))
                                                        
# Create ejection fraction groups (adjust the breaks as needed)
dataset_new$ejection_fraction_group <- cut(dataset_new$ejection_fraction, breaks = c(0, 10, 20, 30, 40, 50, 70, 80))

# Create serum creatinine groups (adjust the breaks as needed)
dataset_new$serum_creatinine_group <- cut(dataset_new$serum_creatinine, breaks = c(0, 2, 4, 6, 8, 10))

# Create bar charts for the grouped variables
bar_chart_age <- ggplot(data = dataset_new, aes(x = age_group, fill = factor(DEATH_EVENT))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Age vs. Death Event",
    x = "Age Group",
    y = "Count",
    fill = "Death Event"
  ) +
  scale_fill_manual(values = c("blue", "red"))  # Switch colors

bar_chart_ejection_fraction <- ggplot(data = dataset_new, aes(x = ejection_fraction_group, fill = factor(DEATH_EVENT))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Ejection Fraction vs. Death Event",
    x = "Ejection Fraction Group",
    y = "Count",
    fill = "Death Event"
  ) +
  scale_fill_manual(values = c("blue", "red"))  # Switch colors

bar_chart_serum_creatinine <- ggplot(data = dataset_new, aes(x = serum_creatinine_group, fill = factor(DEATH_EVENT))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Serum Creatinine vs. Death Event",
    x = "Serum Creatinine Group",
    y = "Count",
    fill = "Death Event"
  ) +
  scale_fill_manual(values = c("blue", "red"))  # Switch colors

# Show the bar charts
print(bar_chart_age)
print(bar_chart_ejection_fraction)
print(bar_chart_serum_creatinine)

```

```{r}
# Load necessary libraries
library(ggplot2)
library(heatmaply)
library(reshape2)

# Calculate the correlation matrix
correlation_matrix <- cor(dataset_new[, c("age", "ejection_fraction", "serum_creatinine", "serum_sodium")])

# Load the necessary libraries
library(ggplot2)
library(reshape2)

# Melt the correlation matrix
correlation_melted <- melt(correlation_matrix)

# Create a heatmap using ggplot2
ggplot(data = correlation_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "black") +
  labs(
    title = "Correlation Heatmap",
    x = "Variables",
    y = "Variables"
  )



```

```{r}
# Load necessary libraries
library(ggplot2)

# Create box plots for numerical variables
boxplot_age <- ggplot(dataset_new, aes(x = factor(DEATH_EVENT), y = age, fill = factor(DEATH_EVENT))) +
  geom_boxplot() +scale_fill_manual(values = c("blue", "red"))+
  labs(
    title = "Box Plot of Age by Death Event",
    x = "Death Event",
    y = "Age",
    fill = "Death Event"
  )

boxplot_ejection_fraction <- ggplot(dataset_new, aes(x = factor(DEATH_EVENT), y = ejection_fraction, fill = factor(DEATH_EVENT))) +
  geom_boxplot() +scale_fill_manual(values = c("blue", "red"))+
  labs(
    title = "Box Plot of Ejection Fraction by Death Event",
    x = "Death Event",
    y = "Ejection Fraction",
    fill = "Death Event"
  )

boxplot_serum_creatinine <- ggplot(dataset_new, aes(x = factor(DEATH_EVENT), y = serum_creatinine, fill = factor(DEATH_EVENT))) +
  geom_boxplot() +scale_fill_manual(values = c("blue", "red"))+
  labs(
    title = "Box Plot of Serum Creatinine by Death Event",
    x = "Death Event",
    y = "Serum Creatinine",
    fill = "Death Event"
  )

# Show the box plots
print(boxplot_age)
print(boxplot_ejection_fraction)
print(boxplot_serum_creatinine)




```

## 5.Future Considerations

Transforming categorical variables like smoking, diabetes, and blood pressure into more granular numerical values could reveal more nuanced correlations with death events. Here's how you might consider converting them:

Smoking: Instead of a binary "Yes/No" for smoking, you could quantify this as "Years of Smoking," which would represent the total number of years an individual has been smoking. This provides a continuous variable that reflects both the duration and potential intensity of smoking habits.

Diabetes: For diabetes, a possible numerical transformation could be "Diabetes Control Score," which could be based on a composite measure of glycemic control (like HbA1c levels).

By quantifying these variables, we might be able to better capture their effects on the risk of death events and potentially uncover patterns that are not visible with categorical data. It's essential to ensure that these numerical values are clinically meaningful and based on actual health records for accurate analysis.
